---
title: "Lake Trout Length-Weight Relationship"
author: "Matt Tyers"
date: "2024-05-01"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning=FALSE, message=FALSE, 
                      fig.width = 9, fig.height=7)
```

```{r}
library(tidyverse)
library(jagsUI)
library(jagshelper)
```


## Length-Weight Relationship for Each Lake

### Model

A log-log regression was used to model weight as a function of length, expanding somewhat on the form used by Lester, et al.  Since fish-level data (lengths and weights) were available from several lakes of differing habitat, the length-weight relationships were fit separately for each lake, with slope and intercept parameters modeled hierarchically as shown below.

$log(W_{j[i]}) = \beta_{0[j]} + \beta_{1[j]}log(L_{j[i]}) + \epsilon_i$

which is equivalent to

$W = e^{\beta_{0[j]}} \times L^{\beta_{1[j]}} \times e^{\epsilon_i}$

in which

$\beta_{0[j]} \sim N(\mu_{\beta_0}, \sigma_{\beta_0}) \text{ for } j \in 1...n_{lakes}$

$\beta_{1[j]} \sim N(\mu_{\beta_1}, \sigma_{\beta_1}) \text{ for } j \in 1...n_{lakes}$

$\epsilon_i \sim N(0,\sigma_{\epsilon})$

### Results

```{r, fig.show='hide'}
# loading all the data, but hiding the data exploration plots, etc
source("R/1_laketrout_lwdata.R")
```


## Asymptotic Length for Each Lake

## Appendix: notes on data and modeling

### Length-Weight Relationship

#### Data

#### Modeling

```{r}

```

