---
title: "An Integrated Model for Estimating Asymptotic Size of Lake Trout"
author: "Matt Tyers"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi=300)
```

```{r LoadSample}
# loading data to report sample sizes
load(file="Rdata/laketrout_sampling_formodel.Rdata")
library(magrittr)
library(jagshelper)

roundn <- paste0(formatC(round(nrow(laketrout), digits=-4), 
                         big.mark=",", format="f", digits=0), 
                 "+")
nw <- sum(!is.na(laketrout$Weight_g)) %>% 
  formatC(big.mark=",", format="f", digits=0)
nl <- sum(!is.na(laketrout$ForkLength_mm)) %>% 
  formatC(big.mark=",", format="f", digits=0)
na <- sum(!is.na(laketrout$Age)) %>% 
  formatC(big.mark=",", format="f", digits=0)

nlw <- sum(!is.na(laketrout$ForkLength_mm & laketrout$Weight_g)) %>% 
  formatC(big.mark=",", format="f", digits=0)
nla <- sum(!is.na(laketrout$ForkLength_mm & laketrout$Age)) %>% 
  formatC(big.mark=",", format="f", digits=0)

nllw <- length(unique(laketrout$LakeName[!is.na(laketrout$ForkLength_mm & laketrout$Weight_g)]))
nlla <- length(unique(laketrout$LakeName[!is.na(laketrout$ForkLength_mm & laketrout$Age)]))

nlakes <- length(unique(laketrout$LakeName))
```

```{r LoadModelResults}
## loading the JAGS output itself (this is a big file)
load(file="interim_posts/int_Winf_modelrun.Rdata")

# ^^^^^ update this file with non-interim model run ^^^^^

```




```{r make_nmat}
ff <- function(x, n=nrow(laketrout_Winf)) { # f for fill
  c(x, rep(NA, n - length(x)))
}

smallernames <- lakenames %>%
  strsplit(split = " Lake") %>%
  sapply("[", 1)
smallernames[lakenames == "Caribou Lake (Cantwell)"] <- "Caribou (Cantwell)"
smallernames[lakenames == "Caribou Lake (Lake Louise)"] <- "Caribou (Lake Louise)"
smallernames[lakenames == "Itkillik lake"] <- "Itkillik"
smallernames[lakenames == "Little Lake Louise"] <- "Little Louise"
smallernames[lakenames == "Sevenmile Lake (Denali Hwy)"] <- "Sevenmile (Denali Hwy)"

nL <- tapply(!is.na(int_Winf_data$L), int_Winf_data$lake, sum)
nW <- tapply(!is.na(int_Winf_data$logW), int_Winf_data$lake, sum)
nA <- tapply(!is.na(int_Winf_data$Age), int_Winf_data$lake, sum)
nLW <- tapply(!is.na(int_Winf_data$L) & !is.na(int_Winf_data$logW), int_Winf_data$lake, sum)
nLA <- tapply(!is.na(int_Winf_data$L) & !is.na(int_Winf_data$Age), int_Winf_data$lake, sum)
nproj <- tapply(laketrout$ProjectTitle, laketrout$LakeNum, \(x) length(unique(x)))

nothing <- "-"
nmat <- data.frame(Lake=smallernames, 
                   Estimate_Winf = ifelse(morphometry$make_estimates, "Yes", "No"),
                   Use_Samples = ifelse(morphometry$use_fish, "Yes", "No"),
                   n_Projects=nothing,
                   n_Length=nothing, n_Weight=nothing, n_Age=nothing, 
                   n_LengthWeight=nothing, n_LengthAge=nothing)
fillnothing <- \(x) ifelse(x > 0, format(x, big.mark=","), nothing)
nmat$n_Length[as.numeric(names(nL))] <- fillnothing(nL)
nmat$n_Weight[as.numeric(names(nW))] <- fillnothing(nW)
nmat$n_Age[as.numeric(names(nA))] <- fillnothing(nA)
nmat$n_LengthWeight[as.numeric(names(nLW))] <- fillnothing(nLW)
nmat$n_LengthAge[as.numeric(names(nLA))] <- fillnothing(nLA)
nmat$n_Projects[as.numeric(names(nproj))] <- fillnothing(nproj)
```

## Abstract


## Purpose (?)


## Introduction

Lake trout are present as a resident species in Alaska, and are long-lived and slow-growing.  In order to effectively manage localized harvest limits, a productivity model must be employed in order to estimate surplus production in a given waterbody, as environmental variability results in variability in productivity at the local scale.  

Following direction in the Wild Lake Trout Management Plan for the Upper Copper-Upper Susitna area lakes, adopted by the Alaska Board of Fisheries (BOF) in 2005 (5AAC 52.060), yield potential (YP) is to be estimated using the lake area (LA) model developed by Evans et al. (1991). In this case YP is the predicted biomass that can be sustainably removed from the lake. This model uses surface area as a measure of available preferred habitat.  As lakes increase in size, so generally does their depth and optimal thermal habitat, where neither temperature nor oxygen is limiting (Marshall 1996).  This model was developed from data compiled from 43 lakes located in Ontario, Canada, and can be expressed as the following:

$$log_{10}(\hat{YP}_{kg}) = 0.60 + 0.72log_{10}(A)$$

in which $\hat{YP}_{kg}$ represents the yield potential expressed in kilograms biomass, and *A* represents lake area measured in hectares.

A newer, more comprehensive model was recently developed which considers information from 456 lakes throughout Canada including the Yukon Territories and Northwest Territories (Lester et al. 2021). Unlike the LA model, this model (hereafter referred to as the Lester model) estimates maximum sustained yield (MSY), which is much different from YP, and of more use to fishery managers. The model incorporates a variety of metrics including not only surface area, but also mean depth, maximum depth, and mean annual air temperature at the location of the lake. In addition, the model allows for the input of other habitat suitability variables such as the proportion of a lake in the hypolimnetic zone, along with population specific metrics such as natural mortality and asymptotic weight ($W_{\infty}$). 

First, the depth ratio (DR) is calculated using:

$$DR = \frac{D_{max}}{D_{mn}}$$ 

where:
	
$D_{max}$ = Maximum depth from bathymetric mapping (m)
	
$D_{mn}$ = Mean depth from bathymetric mapping (m)

Thermocline depth ($D_{th}$) is estimated using:

$$D_{th} = 3.26 \times A^{0.109}  \times D_{mn}^{0.213} \times e^{-0.0263 T}$$

where: 
	
A = surface area (ha)
	
T = mean annual air temperature (°C)

The proportion of the lake volume in the hypolimnetic zone ($pV_{hy}$) is estimated using:

$$pV_{hy}=\left(1 - \frac{D_{th}}{D_{max}}\right)^{DR}$$

The volume of the epibenthic zone ($pV_{eb}$) is estimated using:

$$pV_{eb} = e^{-4.63pV_{hy}}$$

Habitat suitability (S) is summarized using:

$$S = \frac{1}{1 + e^{2.47 + 0.386T - 16.8pV_{hy} }}$$	

Asymptotic length ($L_{\infty}$) is estimated within the Lester model using:

$$L_{\infty} = 957 \times \left(1 - e^{-0.14 \times \left(1 + ln(A)\right)}\right)$$

Asymptotic weight ($W_{\infty}$) is estimated within the Lester Model using:

$$W_{\infty} = \left(\frac{L_{\infty}}{451}\right)^{3.2}$$

An alternative method for estimating asymptotic weight (W_∞) requires taking the mean weight of the heaviest 10% of the fish weighed from the March and June sampling (Lester et al. 2021). This lake specific result will be compared to, and possibly substituted for the results of equation 14.

Biomass density when a population is exploited at MSY ($B_{msy}$) will then be estimated using:

$$B_{msy} = \frac{8.47 \times D_{mn} \times pV_{eb}  × S }{W_{\infty}^{1.33}}$$

Natural mortality rate of the population (M) is estimated using:

$$M=  \frac{0.26 e^{0.021T + 0.0004T^2}}{W_{\infty}^{0.30}} $$

MSY per hectare surface area ($MSY_{ha}$) is estimated using:

$$MSY_{ha} = B_{msy} \times M$$

MSY expressed in kilograms biomass is estimated using:

$$MSY = MSY_{ha} \times A$$                                                             


Application of the Lester Model to lake trout in Alaska lakes may be substantially improved by direct estimation of asymptotic weight $W_{\infty}$ when possible.  Since estimated *MSY* is inversely proportional to asymptotic weight raised to a power, estimates of *MSY* are therefore particularly mathematically sensitive to how asymptotic weight is estimated.  In the computational framework presented by Lester, et al., $W_{\infty}$ is estimated by means of a sequence of relationships: first, the relationship observed between lake area and asymptotic length observed in Canada; then, the aggregate relationship between length and weight observed in Canada.

The Alaska Lake Database (ALDAT, 2013-) serves as a compilation of sampling data from numerous ADF&G research projects and sampling efforts.  Not only does this database provides data at a much finer resolution, the information provided is much more directly applicable to management of the lakes represented, or those closer in proximity and similarity than those used by Lester, et al.  For `r nlakes` lakes, there is sampling data on `r roundn` individual lake trout (`r nw` weights; `r nl` lengths; `r na` ages), which can be used to gain inference on pairwise relationships between variables (`r nlw` paired length x weight observations from `r nllw` lakes; `r nla` paired length x age observations from `r nlla` lakes).  Not only can the relationships between length and weight and between lake area and asymptotic length be estimated for Alaska, they can be estimated specifically for each of the `r nlakes` lakes and applied to additional lakes as appropriate, and parameters reported by Lester, et. al. may be incorporated as priors within a Bayesian modeling framework.

## Objectives



## Methods

### Available Data

Sampling data were used from `r sum(as.numeric(nmat$n_Projects), na.rm=TRUE)` unique sampling efforts, across `r sum(morphometry$use_fish)` lakes deemed to have sufficiently representative and high-quality data.  Out of `r nrow(morphometry)` lakes of interest with lake trout present as a resident species, estimation of asymptotic size is desired by management for `r sum(morphometry$make_estimates)` lakes, `r sum(morphometry$make_estimates & morphometry$use_fish)` of which may make direct use of sampling data and `r sum(morphometry$make_estimates & !morphometry$use_fish)` of which have not yet been sampled.  After filtering to the most reliable data (Appendix AAA), a wealth of data remain, and are summarized for each lake in **Table 1**.

```{r}
knitr::kable(nmat,
             align = c(rep("l", 3), rep("r", ncol(nmat)-3)))
```

### The Integrated Model

Implementation of an integrated model as outlined below allows simultaneous use of all available information, as well as appropriate propagation of all uncertainty in estimation.  The model can be conceptualized as consisting of six inter-related components:

#### Length ~ Age component

The relationship between length and age was modeled using a Von Bertalanffy (LVB) growth function, with multiplicative (lognormal) error.  For fish *i* within lake *j*:

$L[j[i]] \sim LogN(log(\mu_{Lt}[j[i]]), \sigma_{Lt})$

$\mu_{Lt}[j[i]] = L_{\infty}[j](1-e^{-k[j](t[j[i]]-t_0[j])})$

Growth parameters $L_{\infty}$, $k$, and $t_0$ were modeled separately for each lake *j*, with hierarchical distributions according to the form below.  Note that the hierarchical mean $t_0$ is mildly constrained to be near zero; this is to aid convergence and to ensure that it falls within a biologically reasonable range.

$t_0[j] \sim N(\mu_{t_0}, \sigma_{t_0})$

$\mu_{t_0} \sim N(0,1)$

$\sigma_{t_0} \sim Diffuse Unif$

$k[j] \sim N(\mu_{k}, \sigma_{k})$

$\mu_{k} \sim Diffuse Norm$

$\sigma_{k} \sim Diffuse Unif$


#### $L_{\infty}$ ~ Area component

Growth parameter $L_{\infty}$ may be thought of as hierarchically distributed, but with a central trend defined not by a global mean but according to the relationship with lake area reported by Lester, et al., and with multiplicative (lognormal) process error.  Parameters $\gamma$ and $\lambda$ were given weakly informative priors centered on the values reported by Lester, et al., and with coefficients of variation given below.  For lake *j*:

$L_{\infty}[j] \sim LogN(log(\mu_{LA}[j]), \sigma_{LA})$

$\mu_{LA}[j] = \gamma(1-e^{-\lambda(1+log(A[j]))})$

$\sigma_{LA} \sim Diffuse Unif$

$\gamma \sim N(\gamma_{Lester}, 100\% \times \gamma_{Lester})$

$\lambda \sim N(\lambda_{Lester}, 250\% \times \lambda_{Lester})$

#### Weight ~ Length component

A log-log regression was used to model weight as a function of length, expanding somewhat on the form used by Lester, et al.  Since fish-level data (lengths and weights) were available from several lakes of differing habitat, the length-weight relationships were estimated separately for each lake, with slope and intercept parameters modeled hierarchically as further described.  For fish *i* within lake *j*: 

THIS WHOLE DARN PIECE WILL PROBABLY NEED TO BE REFINED AFTER MODEL SELECTION

$log(W\left[j[i]\right]) \sim N \left(\mu_{WL} [j[i]], \sigma_{WL}\right)$

$\mu_{WL} [j[i]] = \beta_0[j] + \beta_1[j] log(L[j[i]])$

$\sigma_{WL} \sim Diffuse Unif$

$\beta_0[j] \sim N(\mu_{\beta_0}, \sigma_{\beta_0})$

$\mu_{\beta_0} \sim Diffuse Norm$

$\sigma_{\beta_0} \sim Diffuse Unif$

$\beta_1[j] \sim N(\mu_{\beta_1}, \sigma_{\beta_1})$

$\mu_{\beta_1} \sim Diffuse Norm$

$\sigma_{\beta_1} \sim Diffuse Unif$



#### Length Quantile ~ $L_{\infty}$ component

In the absence of sufficient paired age and length data, it is common practice to use a quantile (percentile) value from a length sample as a proxy for asymptotic length.  Given the relative wealth of length samples within our dataset, it is certainly worth incorporating this information into an integrated model.

Length quantile values for each lake *j* were treated as observations of asymptotic length, with additive (normal) error.  The error standard deviation was conceptualized as consisting of two independent components: the uncertainty due to estimation of a population length quantile, which will be approximately proportional to the inverse square root of the sample size, and the uncertainty within the relationship between the population length quantile and asymptotic length.  

The model was run for a sequence of trial percentile values in sequence, ranging from the 85th to 99th percentiles, and data values (length percentiles) were compared to the respective posterior predictive distributions.  For this dataset, using the 95th percentile to estimate asymptotic length resulted in the best agreement between data values and posterior predictive distributions, which can be taken as evidence of consistency among all available sources of information.

$q_L[j] \sim N(L_{\infty}, \sigma_L[j])$

$\sigma_L[j] = \sqrt{\eta_L^2 + \frac{\zeta_L^2}{n_L[j]}}$

$\eta_L \sim Diffuse Unif$

$\zeta_L \sim Diffuse Unif$


#### Weight Quantile ~ $W_{\infty}$ component

Similarly, weight quantiles were treated as observations of asymptotic weight as available, with estimable error according to the same form.  The relationship between length and weight is non-linear; however, quantile values are invariant to monotonic transformations.  Therefore, the 95th percentile values of weight were used.

$q_W[j] \sim N(W_{\infty}, \sigma_W[j])$

$\sigma_W[j] = \sqrt{\eta_W^2 + \frac{\zeta_W^2}{n_W[j]}}$

$\eta_W \sim Diffuse Unif$

$\zeta_W \sim Diffuse Unif$


#### $W_{\infty} \sim L_{\infty}$ relationship

Finally, the relationship between asymptotic length and asymptotic weight for lake *j* was treated as deterministic (that is, without additional observation error), according to the log-log regression parameters estimated for that lake.  Since asymptotic size functions as a central trend that is not subject to fish-level variability, the only variability that is necessary to incorporate is that due to estimation of the regression parameters themselves.  Inclusion of this step in the same model that is also estimating the length-weight relationship will therefore propagate this uncertainty.

$log(W_{\infty}[j]) = \beta_0[j] + \beta_1[j] log(L_{\infty}[j])$


### Model fitting and diagnostics

Model fitting was performed using JAGS (Just Another Gibbs Sampler, Plummer 2003) called through R^[`r capture.output(print(citation(), style="text"))`] using the jagsUI^[`r capture.output(print(citation("jagsUI"), style="text"))`] package, with output manipulation, convergence diagnostics, and plotting performed using the jagshelper package^[`r capture.output(print(citation("jagshelper"), style="text"))`]. `r english::Words(ncores)` chains of `r formatC(niter, big.mark=",", format="f", digits=0)` MCMC samples were taken, with the first half discarded as burn-in and the remainder thinned to 1,000 samples per parameter node per chain.  Convergence was assessed graphically by means of trace plots and corroborated using the Gelman-Rubin convergence diagnostic (R hat, Gelman & Rubin 1992).  R hat values for all parameter nodes were well within a threshold of 1.1, and all those for global hyperparameters or parameter nodes pertaining to Crosswind Lake were within a threshold of 1.01.   **UPDATE THIS AS NEEDED**

Model appropriateness with respect to overparameterization was further assessed by means of posterior predictive quantile-quantile plots whenever possible **list specific cases?**.  Between-parameter correlations were estimated and bivariate plots were constructed as appropriate, in order to ensure that no additional reparameterization was necessary to facilitate convergence **klunky and maybe not needed**.


## Results

```{r}
# A good-enough sideways version of caterpillar (data is on x-axis)
side_cat2 <- function(x, xnames, col=1, medsort=FALSE, x0=FALSE, highlight=NA, xlab="", ...) { 

  # defining colors for plotting (highlight vs not)
  if(all(!is.na(highlight))) {
    col <- ifelse(highlight, col, adjustcolor(col, alpha.f=.2))
  } else {
    col <- rep(col, length(xnames))
  }
  
  # subsetting input matrix (only non-NA columns)
  notNA <- apply(x, 2, \(xx) all(!is.na(xx)))
  x <- x[, notNA]
  xnames <- xnames[notNA]
  col <- col[notNA]
  
  # defining interval types and calculating
  ci <- c(0.5, 0.95)
  loq <- apply(x, 2, quantile, p = (1 - ci)/2, na.rm = T)
  hiq <- apply(x, 2, quantile, p = 1 - (1 - ci)/2, na.rm = T)
  meds <- apply(x, 2, median, na.rm=T)
  
  # reordering if we are sorting by post median
  if(!medsort) {
    yplot <- seq_along(xnames)
  } else {
    yplot <- rank(meds)
  }

  # defining xlims 
  if(x0) {
    xlims <- range(0, loq, hiq, na.rm=TRUE)
  } else {
    xlims <- range(loq, hiq, na.rm=TRUE)
  }
  
  # actually plotting
  plot(x=meds, y=yplot,
       xlim=xlims, ylim=c(ncol(x),1),
       pch="|", col=col,
       ylab="", xlab=xlab, yaxt="n", ...=...)
  
  axis(side=2, at=yplot, labels=xnames, las=2, cex.axis=1)
  grid(ny=NA)
  
  segments(x0=loq[1,], x1=hiq[1,], y0=yplot,
           lend=1, lwd=3, col=col)
  segments(x0=loq[2,], x1=hiq[2,], y0=yplot,
           lend=1, lwd=1, col=col)
}

# A good-enough upright version of caterpillar (data is on y-axis)
# this is totally ridiculous but I got tired of forcing caterpillar to behave
y_cat2 <- function(x, xnames=NA, xval=NA, col=1, medsort=FALSE, x0=FALSE, highlight=NA,  ...) { 

  # defining colors for plotting (highlight vs not)
  if(all(!is.na(highlight))) {
    col <- ifelse(highlight, col, adjustcolor(col, alpha.f=.2))
  } else {
    col <- rep(col, ncol(x))
  }
  
  # subsetting input matrix (only non-NA columns)
  notNA <- apply(x, 2, \(xx) all(!is.na(xx)))
  x <- x[, notNA]
  xnames <- xnames[notNA]
  xval <- xval[notNA]
  col <- col[notNA]
  
  # defining interval types and calculating
  ci <- c(0.5, 0.95)
  loq <- apply(x, 2, quantile, p = (1 - ci)/2, na.rm = T)
  hiq <- apply(x, 2, quantile, p = 1 - (1 - ci)/2, na.rm = T)
  meds <- apply(x, 2, median, na.rm=T)
  
  if(!all(is.na(xnames))) {
  # reordering if we are sorting by post median
  if(!medsort) {
    yplot <- 1:ncol(x)
  } else {
    yplot <- rank(meds)
  }
  }
  if(!all(is.na(xval))) yplot <- xval

  # defining YYYlims 
  if(x0) {
    xlims <- range(0, loq, hiq, na.rm=TRUE)
  } else {
    xlims <- range(loq, hiq, na.rm=TRUE)
  }
  
  # actually plotting
  plot(y=meds, x=yplot,
       ylim=xlims,#, xlim=c(1,ncol(x)),
       pch="-", col=col,
       # xlab="", ylab=xlab, #xaxt="n", 
  # )
       ...=...)
  
  if(!all(is.na(xnames))) {
    axis(side=1, at=yplot, labels=xnames, las=2, cex.axis=0.8)
  }
  grid(nx=NA, ny=NULL)
  
  segments(y0=loq[1,], y1=hiq[1,], x0=yplot,
           lend=1, lwd=3, col=col)
  segments(y0=loq[2,], y1=hiq[2,], x0=yplot,
           lend=1, lwd=1, col=col)
}


parmar <- par("mar")
```

Plots of param ests for all lakes:
b0, b1, k, t0

Plots of sigW & sigL wrt sample sizes

Plots of Linf & Winf wrt lake area (with lester line)

BIG DARN TABLE with ests & SE per lake


#### Length ~ Age component

The relationship between length and age was able to be directly estimated for `r sum(!is.na(int_Winf_jags_out$q50$k))` lakes, directly informing estimation of the asymptotic length parameter $L_{\infty}$.  Additional LVB parameters *k* and *t0* were modeled hierarchically, with inferences shown in Figures BBB and CCC.  **Make some comparison to priors used?? Perhaps include this as a model parameter**

```{r, fig.width=8, fig.height=5, fig.cap="LVB growth parameter k for lakes with sufficient age data.  Light and heavy bars correspond to 95% and 50% credible intervals, respectively, with black bars representing lakes for which estimates of asymptotic size are required."}
par(mar=c(5, 9, 1, 1)+0.1)
par(family="serif")

side_cat2(x=int_Winf_jags_out$sims.list$k,
          xnames=smallernames, 
          medsort=FALSE,
          highlight = morphometry$make_estimates,
          xlab="k")
```

```{r, fig.width=8, fig.height=5, fig.cap="LVB growth parameter t0 for lakes with sufficient age data.  Light and heavy bars correspond to 95% and 50% credible intervals, respectively, with black bars representing lakes for which estimates of asymptotic size are required."}
parmar <- par("mar")
par(mar=c(5, 9, 1, 1)+0.1)
par(family="serif")

side_cat2(x=int_Winf_jags_out$sims.list$t0,
          xnames=smallernames, 
          medsort=FALSE,
          highlight = morphometry$make_estimates,
          xlab="t0 (years)")
```



#### $L_{\infty}$ ~ Area component

Asymptotic length $L_{\infty}$ was modeled relatively precisely for lakes with paired length and age samples, and the relationship between asymptotic length and lake area was used to estimate $L_{\infty}$ for the remainder of the lakes (Figure DDD).  We estimate an increasing relationship between asymptotic length and lake area, though not as strong as that reported by Lester, et al., and with asymptotic length estimated slightly larger for small lakes.  There is much larger estimation variance associated with lakes with no paired length and age samples; however, this may be expected due to the similar degree of spread in asymptotic length for lakes with paired length and age data.


```{r, fig.width=8, fig.height=7, fig.cap="Estimated asymptotic length for all lakes.  Light and heavy vertical bars correspond to 95% and 50% credible intervals, respectively, with black bars representing lakes for which estimates of asymptotic size are required.  The dashed curve corresponds to the relationship reported by Lester, et al."}

par(mar=c(5, 5, 1, 1)+0.1)
par(family="serif")

y_cat2(int_Winf_jags_out$sims.list$Linf,
       xval=morphometry$SurfaceArea_h, 
       highlight=morphometry$make_estimates, 
       x0=TRUE,
       log="x", 
       xlab="Surface area (ha)",
       ylab="Asymptotic length (mm FL)")
curve(int_Winf_data$gam_lester * (1 - exp(-int_Winf_data$lam_lester * (1 + log(x)))), add=TRUE, lty=3)


# par(mar=parmar)
```



#### Weight ~ Length component

Similarly, the relationship between length and weight was estimated relatively precisely for lakes with paired length and weight data, and the hierarchical means and variances of regression parameters was propagated to estimate the regression parameters of lakes with no paired length and weight data.  Regression parameters *b0* and *b1* were generally very similar to those reported by Lester et al. (Figure EEE), though most lakes with paired data exhibited strong enough differences from these values to illustrate the value of estimating this relationship separately for each lake.

```{r, fig.width=8, fig.height=10, fig.cap="Log(weight) ~ log(length) regression parameter b0.  Light and heavy horizontal bars correspond to 95% and 50% credible intervals, respectively, with black bars representing lakes for which estimates of asymptotic size are required.  The vertical dotted line corresponds to the global value used by Lester, et al."}
par(mar=c(5, 9, 1, 1)+0.1)
par(family="serif")

side_cat2(x=int_Winf_jags_out$sims.list$b0_interp,
          xnames=smallernames, 
          medsort=FALSE,
          highlight = morphometry$make_estimates,
          xlab="b0")
abline(v=log(1/451^3.2), lty=3)
```

```{r, fig.width=8, fig.height=10, fig.cap="Log(weight) ~ log(length) regression parameter b1.  Light and heavy horizontal bars correspond to 95% and 50% credible intervals, respectively, with black bars representing lakes for which estimates of asymptotic size are required.  The vertical dotted line corresponds to the global value used by Lester, et al."}
par(mar=c(5, 9, 1, 1)+0.1)
par(family="serif")

side_cat2(x=int_Winf_jags_out$sims.list$b1,
          xnames=smallernames, 
          medsort=FALSE,
          highlight = morphometry$make_estimates,
          xlab="b1")
abline(v=3.2, lty=3)
```



#### Length Quantile ~ $L_{\infty}$ and Weight Quantile ~ $W_{\infty}$ components

The relationships between sample quantiles (length and weight) and the respective asymptotic values ($L_{\infty}$ and $W_{\infty}$) were increasingly precise as sample size increased.  The relationship between the relationship standard deviations and the numbers of length or weight samples per lake is illustrated in Figure FFF and GGG.  In the case of length, the relationship standard deviation nearly approaches an asymptotic value, as a result of the comparative wealth of length samples in many lakes.

```{r, fig.width=8, fig.height=7, fig.cap="Estimated standard deviation parameter relating length sample quantile to asymptotic length, as a function of number of length samples per lake.  The light and dark bands correspond to 95% and 50% credible envelopes, and the solid line corresponds to posterior median."}
par(mar=c(5, 5, 1, 1)+0.1)
par(family="serif")

envelope(int_Winf_jags_out$sims.list$sig_L[, int_Winf_data$whichlakes_L],
         x=int_Winf_data$nL[int_Winf_data$whichlakes_L], 
         log="x",
         xlab="Number of length samples per lake", 
         ylab="sig_L (mm FL)",
         col=1,
         ylim=c(0, max(int_Winf_jags_out$q97.5$sig_L, na.rm=TRUE)))
```


```{r, fig.width=8, fig.height=7, fig.cap="Estimated standard deviation parameter relating weight sample quantile to asymptotic weight, as a function of number of weight samples per lake.  The light and dark bands correspond to 95% and 50% credible envelopes, and the solid line corresponds to posterior median."}
par(mar=c(5, 5, 1, 1)+0.1)
par(family="serif")

envelope(int_Winf_jags_out$sims.list$sig_W[, int_Winf_data$whichlakes_W],
         x=int_Winf_data$nW[int_Winf_data$whichlakes_W], 
         log="x",
         xlab="Number of weight samples per lake", 
         ylab="sig_W (kg)",
         col=1,
         ylim=c(0, max(int_Winf_jags_out$q97.5$sig_W, na.rm=TRUE)))
```



#### $W_{\infty} \sim L_{\infty}$ relationship

While estimation of $W_{\infty}$ is not purely a result of propagating estimates of $L_{\infty}$ through the length-weight relationship, it is illustrative to present the relationship between asymptotic weight and lake area, particularly because this is how asymptotic weight is estimated by Lester, et al.  While a loose relationship is indeed apparent, there also exists a strong degree of variability between lakes, further illustrating the utility of lake-specific estimation. 

```{r, fig.width=8, fig.height=7, fig.cap="Estimated asymptotic weight for all lakes.  Light and heavy vertical bars correspond to 95% and 50% credible intervals, respectively, with black bars representing lakes for which estimates of asymptotic size are required.  The dashed curve corresponds to the relationship reported by Lester, et al."}

par(mar=c(5, 5, 1, 1)+0.1)
par(family="serif")

y_cat2(x=int_Winf_jags_out$sims.list$Winf,
       xval=morphometry$SurfaceArea_h, 
       highlight=morphometry$make_estimates, 
       x0=TRUE,
       log="x", 
       xlab="Surface area (ha)",
       ylab="Asymptotic weight (kg)")
curve(exp(-19.56 +
            3.2*log(int_Winf_data$gam_lester * (1 - exp(-int_Winf_data$lam_lester * (1 + log(x)))))),
      add=TRUE, lty=3)


# par(mar=parmar)
```




#### $L_{\infty} and  W_{\infty}$ for all lakes

Asymptotic length and weight is presented below for all lakes (Figures HHH and III), showing a wide range in asymptotic sizes and particularly precision in estimation.

```{r, fig.width=8, fig.height=11, fig.cap="Estimated asymptotic length for all lakes.  Light and heavy horizontal bars correspond to 95% and 50% credible intervals, respectively, with black bars representing lakes for which estimates of asymptotic size are required."}

par(mar=c(5, 9, 1, 1)+0.1)
par(family="serif")

side_cat2(x=int_Winf_jags_out$sims.list$Linf,
          xnames=smallernames, 
          medsort=TRUE, x0=TRUE,
          highlight = morphometry$make_estimates,
          xlab="Asymptotic length (mm FL)")
```

```{r, fig.width=8, fig.height=11, fig.cap="Estimated asymptotic weight for all lakes.  Light and heavy horizontal bars correspond to 95% and 50% credible intervals, respectively, with black bars representing lakes for which estimates of asymptotic size are required."}

par(mar=c(5, 9, 1, 1)+0.1)
par(family="serif")

side_cat2(x=int_Winf_jags_out$sims.list$Winf,
          xnames=smallernames, 
          medsort=TRUE, x0=TRUE,
          highlight = morphometry$make_estimates,
          xlab="Asymptotic weight (kg)")


# par(mar=parmar)
```











## Discussion



## References

Alaska Lake Database (ALDAT).  2013-.  Alaska Department of Fish and Game, Division of Sport Fish.  Available at http://www.adfg.alaska.gov/SF_Lakes/ (accessed December 23, 2025).

Albert in prep winter 22-23

Albert, M. L., and M. R. Chari.  2026. Stock assessment and yield potential of lake trout in the Tangle Lakes system, 2023–2024. Alaska Department of Fish and Game,Fishery Data Series No. XX-XX, Fairbanks.

Burr, J. M. 1995. Lake trout studies in the AYK Region, and burbot index of abundance in Galbraith Lake, 1994.  Alaska Department of Fish and Game, Fisherv Data Series No. 95-30. Anchorage.

Burr, J.M., 1987. Stock assessment and biological characteristics of lake trout populations in interior Alaska, 1986

Burr, J.M., 1988. Stock assessment and biological characteristics of lake trout populations in interior Alaska, 1987

Burr, J.M., 1989. Stock assessment and biological characteristics of lake trout populations in interior Alaska, 1988

Burr, J.M., 1990. Stock assessment and biological characteristics of lake trout populations in interior Alaska, 1989

Burr, J.M., 1991. Lake trout population studies in interior Alaska 1990, including abundance estimates of lake trout in Glacier, Sevenmile, and Paxson Lakes during 1989.

Burr, J.M., 1992.  Studies of lake trout in Sevenmile Lake and the Tangle Lakes during 1991 

Burr, J.M., 1994. Evaluations of introduced lake trout in the Tanana Drainage and population abundance of lake trout in Sevenmile Lake. Alaska Department of Fish and Game, Fishery Data Series No. 94-18, Anchorage.

Evans, D. O., J. M. Casselman, C. C. Wilcox. 1991. Effects of exploitation, loss of nursery habitat, and stocking on the dynamics and productivity of lake trout populations in Ontario lakes. Lake Trout Synthesis. Ontario Ministry of Natural Resources, Toronto.

Gelman, A., & Rubin, D. B. (1992). Inference from Iterative Simulation Using Multiple Sequences. Statistical Science, 7(4), 457–472. http://www.jstor.org/stable/2246093

Jonson, R.R., and K.D. Troyer. 1994.  Survey of lake trout and Arctic char in Walker Lake, Gates of the Arctic National Park and Preserve, 1987-1989. Fish and Wildlife Service, Fairbanks Fishery Resource Office, Alaska Fisheries technical report Number 25, Fairbanks, Alaska.

Lester, N. P., B. J. Shuter, M. L. Jones and S. Sandstrom. 2021. A general, life history-based model for sustainable exploitation of lake charr across their range, pages 429-485 [In] Muir, A. M., C. C. Krueger, M. J. Hansen and S. C. Riley, editors. The lake charr Salvelinus namaycush: biology, ecology, distribution, and management. Fish and Fisheries 39. 

Mansfield, K. in prep. Lake Trout Abundance in Sevenmile Lake, 2023–2024. Alaska Department of Fish and Game, Fishery Data Series No. YY-XX, Anchorage.

Marshall, T. R. 1996. A hierarchal approach to assessing habitat suitability and yield potential of lake trout. Canadian Journal of Fisheries and Aquatic Sciences 53 (Suppl. 1): 332-341.

Parker, J. F., Scanlon, B., and K. Wuttig. 2001. Abundance and Composition of Lake Trout in Fielding (1999) and Island (2000) Lakes, Alaska. Alaska Department of Fish and Game, Fishery Data Series No. 01-31, Anchorage.

Schwanke, C. J., and M. L. Albert. 2019. Estimation of abundance and yield potential of lake trout in Chandler Lake, 2017–2018. Alaska Department of Fish and Game, Fishery Data Series No. 19-30, Anchorage.

Schwanke, Corey J. 2013. Stock assessment of lake trout in Fielding Lake, 2010-2011. Alaska Department of Fish and Game, Fishery Data Series No. 13-55, Anchorage.

Szarzi, N, and D. B. Bernard. 1997. Evaluation of lake trout stock status and abundance in selected lakes in the Upper Copper and Upper Susitna Drainages, 1995. Alaska Department of Fish and Game, Fishery Data Series No. 97-5, Anchorage.

Taube, T. T. 1996. Lake Trout Studies in the AYK Region, 1995. Alaska Department of Fish and Game, Fishery Manuscript No. 96-3, Anchorage.

Taube, T. T. 1997. Lake Trout Studies in the AYK Region, 1996. Alaska Department of Fish and Game, Fishery Manuscript Number 97-2, Anchorage.

Taube, T. T., K. Wuttig, and L. Stuby. 1998. Lake Trout Studies in the AYK Region, 1997. Alaska Department of Fish and Game, Fishery Data Series 98-24, Anchorage.

Troyer, K. D., and R. R. Johnson. 1994. Survey of lake trout and Arctic char in the Chandler Lakes system, Gates of the Arctic National Park and Preserve, 1987 and 1989. U.S. Fish and Wildlife Service, Fairbanks Fishery Resource Office, Alaska Fisheries technical report Number 26, Fairbanks, Alaska.

Wuttig, K. 2010. Stock assessment of lake trout in Paxson Lake, 2002-2004. Alaska Department of Fish and Game, Fishery Data Series No. 10-46, Anchorage.



## Appendix

Data filtration 

candidate W~L models and model selection?

ALL THE PLOTS ALL THE LAKES!
